---
title: "Geospatial Analysis"
author: "Wan Kee"
date: "22 February 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
#  freeze: true
  warning: false
  message: false
  error: true
---

# 1. Learning Objective

A choropleth map is a thematic map composed of **coloured polygons** and is used to represent statistical data using the color mapping symbology technique.

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

-   Plot choropleth maps using tmap package

# 2. Import Packages

```{r}
pacman::p_load(tidyverse, sf, tmap, psych)
```

# 3. Load Data

::: panel-tabset
## Master Plan Subzone

`mpsz` is the Master Plan 2014 Subzone Boundary in ESRI shapefile format. It can be downloaded at data.gov.sg This is a geospatial data. It consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2014. `st_read()` function of sf package to import the MP14_SUBZONE_WEB_PL shapefile into R as a simple feature data frame.

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz
```

## Population

`pop` is the Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format (i.e. respopagesextod2011to2020.csv). This is an aspatial data fie. It can be downloaded at Department of Statistics, Singapore Although it does not contain any coordinates values, but it’s PA and SZ fields can be used as unique identifiers to geocode to MP14_SUBZONE_WEB_PL shapefile.

```{r}
pop <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
glimpse(pop)
```
:::

# 4. Prepare Data

::: panel-tabset
## Recode

The following attributes will be derived:

YOUNG: Age group `0_to_4` and `20_to_24` ECONOMY ACTIVE: Age group `25_to_29` and `60_to_64` AGED: Age group `65_and_above` TOTAL: All age group DEPENDENCY: Ratio between YOUNG and AGED : ECONOMIC ACTIVE

```{r}
pop2020 <- pop %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(POP = sum(Pop)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>%
  mutate(ECONOMY_ACTIVE = rowSums(.[7:11]) + rowSums(.[13:15]))%>%
  mutate(AGED = rowSums(.[16:21])) %>%
  mutate(TOTAL = rowSums(.[3:21])) %>%
  mutate(DEPENDENCY = (YOUNG + AGED)/ECONOMY_ACTIVE) %>%
  select(PA, SZ, YOUNG, ECONOMY_ACTIVE, AGED, TOTAL, DEPENDENCY)
glimpse(pop2020)
```

```{r}
describe(pop2020)
```

## Uppercase

`PA` and `SZ` of `pop2020` and `SUBZONE_N` and `PLN_AREA_N` of `mpsz`are standardized to uppercase.

```{r}
pop2020 <- pop2020 %>%
  mutate_at(.vars = vars(PA, SZ), .funs = funs(toupper)) %>%
  filter(ECONOMY_ACTIVE > 0)
describe(pop2020)
```

## Left Join

`pop2020` and `mpsz` are joined using `SZ` and `SUBZONE_N`.

```{r}
mpszpop2020 <- left_join(mpsz, pop2020,
                         by = c("SUBZONE_N" = "SZ"))
glimpse(mpszpop2020)
```
:::

::: {.callout-note collapse="true"}
# Save Point

```{r}
write_rds(mpszpop2020, "data/mpszpop2020.rds")
```
:::

# 5. Choropleth Map

::: {.callout-note collapse="true"}
## Load Point

```{r}
mpszpop2020 <- readRDS("data/mpszpop2020.rds")
```
:::

## qtm()

`qtm()` plots a quick thematic map and is a convenient wrapper of the main plotting method. The `fill` argument is either a colour or a data variable to draw a choropleth. The disadvantge of `qtm()` makes aesthetics of individual layers harder to control.

```{r}
tmap_mode("plot")
qtm(mpszpop2020, fill = "DEPENDENCY")
```

## tmap()

The basic building block of tmap is `tm_shape()` followed by the layer elements `tm_shape()` to define the input data and `tm_polygons()` to draw the planning subzone polygons.

Note that `tm_polygons()` is a wrapper of `tm_fill()` and `tm_border()`. `tm_fill()` shades the polygons by using the default colour scheme and `tm_borders()` adds the borders of the shapefile onto the choropleth map where the `alpha` argument is used to define transparency number between 0 (totally transparent) and 1 (default; not transparent).

tmap provides a total ten data classification methods, namely: fixed, sd, equal, pretty, quantile, kmeans, hclust, bclust, fisher, and jenks. The default interval binning used to draw the choropleth map is **pretty** and the default colour scheme is YlOrRd of ColorBrewer. By default, Missing value will be shaded in grey.

::: panel-tabset
### pretty

The **pretty** style chooses a number of breaks not necessarily equal to n using pretty, but likely to be legible.

```{r}
tm_shape(mpszpop2020)+
  tm_polygons("DEPENDENCY")
```

### quantile

The **quantile** style provides quantile breaks.

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

### kmeans

The **kmeans** style uses kmeans to generate the breaks; it may be anchored using `set.seed`.

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```

### bclust

The **hclust** style uses hclust to generate the breaks using hierarchical clustering. The **bclust** style uses bclust to generate the breaks using bagged clustering.

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "bclust") +
  tm_borders(alpha = 0.5)
```

### fisher

The **fisher** style uses the algorithm proposed by W. D. Fisher (1958) and discussed by Slocum et al. (2005) as the Fisher-Jenks algorithm. This style will subsample by default for more than 3000 observations.

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "fisher") +
  tm_borders(alpha = 0.5)
```

### jenks

`jenks` is a quantile data classification that used 5 classes. The jenks style has been ported from Jenks' code, and has been checked for consistency with ArcView, ArcGIS, and MapInfo

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```
:::

## Breaks

Breaks are computed internally or set explicitly by means of the breaks argument to the `tm_fill()`. It is a good practice to obtain descriptive statistics on the variable before setting the break points.

```{r}
summary(mpszpop2020$DEPENDENCY)
```

Based on the output, the break points are set at 0.60, 0.70, 0.80, and 0.90. In addition, a minimum and maximum will also be set at 0 and 100. Therefore, the breaks vector is c(0, 0.60, 0.70, 0.80, 0.90, 1.00).

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

## Multiple Choropleth Plots

```{r}
tm_shape(mpszpop2020)+ 
  tm_polygons(c("YOUNG","AGED"),
          style = c("pretty", "pretty"), 
          palette = list("Blues","Purples")) +
  tm_layout(legend.position = c("right", "bottom"))
```

```{r}
tm_shape(mpszpop2020)+ 
  tm_polygons(c("YOUNG","AGED"),
          style = c("quantile", "quantile"), 
          palette = list("Blues","Purples")) +
  tm_layout(legend.position = c("right", "bottom"))
```

## Group-by Variable

```{r}
tm_shape(mpszpop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=FALSE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpszpop2020[mpszpop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

# 6. Geospatial Data Point

Visualizing geospatial point data typically involves plotting points on a map to represent locations or events. Using the mapping library to create a map and add the data points as layers, the map appearance can be customized to enhance visualization, such as background, colors, and labels.

Proportional symbol maps (also known as graduate symbol maps) are a class of maps that use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon.

# 7. Load Data

SGPools

`SGPools` contains the locations of Singapore outlets and branches and their respective winnings.

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
glimpse(sgpools)
```

# 8. Prepare Data

`st_as_sf()` converts sgpools data frame into a simple feature data frame. The `coords` argument defines the x-coordinates and y-coordinates. The `crs` argument specifies the coordinates system in epsg format where EPSG: 3414 is Singapore SVY21 Projected Coordinate System.

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

# 9. Interactive Geospatial Point Map

`tm_bubbles()` adds bubble symbols to the map and are often used to represent point data.

The `size` argument specifies the attribute used to determine the size of the bubbles. It can take on a **numeric value** as a standard size or a **numeric attribute** where the size will be proportional to the values of the "Gp1Gp2 Winnings" attribute.

The `col` argument specifies the color of the bubbles. It can be a **colour value** or a **categorical attribute**.

```{r}
tmap_mode("view")

tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             border.col = "black",
             border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,12))
```

`tm_facets()` divides the map into multiple facets based on the specified attribute. Each unique value of the attribute will result in a separate facet.

The `by` argument specifies the attribute used to create facets. Each unique value of the "OUTLET TYPE" attribute will result in a separate facet.

The `nrow` argument specifies the number of rows in the facet grid. Where nrow = 1, all facets will be arranged in a single row.

The `sync` argument specifies whether the scales of the facets should be synchronized. When set to TRUE, the scales (e.g., zoom levels) of all facets will be synchronized, allowing for easy comparison between facets.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE) +
  tm_view(set.zoom.limits = c(11,12))
```

# 10. Geospatial Analytical Map

Geospatial analytical maps involve the visualization and analysis of spatial data to reveal patterns, trends, and relationships within a geographic context. These maps are designed to provide insights into spatial distributions and variations, allowing for better understanding and interpretation of data. Some common types of geospatial analytical maps include choropleth maps, rate maps, percentile maps, and boxmaps.

# 11. Import Data

`NGA_wp` is a polygon feature data frame providing information on water point of Nigeria at the LGA level.

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

# 12. Choropleth Maps

Choropleth maps use color shading to represent variations in data values across geographic regions. Each region is shaded based on the value of a specific variable, such as population density, average income, or disease prevalence. Choropleth maps are useful for visualizing spatial patterns and disparities in data.

```{r}
tmap_mode("view")

p1 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of functional water point by LGAs",
            legend.outside = FALSE)

p2 <- tm_shape(NGA_wp) +
  tm_fill("total_wp",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of total water point by LGAs",
            legend.outside = FALSE)

tmap_arrange(p2, p1, nrow = 1)
```

# 13. Rate Maps

**Rate maps** are a type of choropleth map that display rates or proportions per unit area. They are commonly used to visualize rates of occurrence or density of phenomena, such as crime rates, disease incidence, or population density. Rate maps often involve normalizing raw counts by a population or area size to allow for meaningful comparisons between different regions.

`mutate()` from dplyr package is used to derive two fields, namely pct_functional and pct_nonfunctional.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)

tm_shape(NGA_wp) +
  tm_fill("pct_functional",
          n = 10,
          style = "equal",
          palette = "Blues",
          legend.hist = TRUE) +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Rate map of functional water point by LGAs",
            legend.outside = TRUE)
```

# 14. Percentile Maps

**Percentile maps** divide the data into equal intervals based on percentiles and assign each interval a unique color or shading. This type of map is useful for identifying relative rankings or distributions of values within a dataset. Percentile maps highlight areas with high or low values relative to the entire dataset and can be effective for comparing distributions across different regions.

The data preparation as below:

::: panel-tabset
## Missing Data

`drop_na()` excludes records with NA.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

## Cumulative Probability

Create customised classification by defining percentile classes where `percent` is c(0,.01,.1,.5,.9,.99,1).

During variables extraction, the geometry is extracted but base R functions cannot deal with the geometry. To prevent an error from `quantile()`, `st_set_geomtry()` is set to NULL to drop geomtry field.

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

## get.var function

We will write an R function `get.var` to extract a variable (wp_nonfunctional) as a vector out of an sf data.frame.

The arguments are `vname` for variable name and `df` as the name of sf data frame It returns `v`, the vector with values.

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```
:::

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}

percentmap("total_wp", NGA_wp)
```

# 15. Boxmaps

Boxmaps, also known as box plot maps, are a spatial representation of box plots. They display summary statistics of a variable (e.g., median, quartiles, outliers) for different geographic areas. Each area is represented by a box plot, with the length of the box indicating the interquartile range (IQR) and the median value shown as a horizontal line within the box. Boxmaps provide a compact visualization of spatial variability and distribution of data.

The data preparation as below:

::: panel-tabset
## boxbreak function

`boxbreak` is an R function that creating break points for a box map. The arguments `v` indicates vector with observations, `multi` is the multiplier for IQR (default 1.5). The function returns `bb`, a vector with 7 break points compute quartile and fences.

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

## get.var function

`get.var` is a function to extract a variable as a vector out of an sf data frame. The arguments `vname` is the variable name and `df` is the name of sf data frame. It returns `v`, a vector with values.

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```
:::

```{r}
boxmap <- function(vnam, df, legtitle=NA,
                   mtitle="Box Map", mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
             labels = c("lower outlier", "< 25%", "25% - 50%", "50% - 75%", "> 75%", "upper outlier")) +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left", "top"))
}

tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
