---
title: "Geospatial Analysis"
author: "Wan Kee"
date: "22 February 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
#  freeze: true
  warning: false
  message: false
  error: true
---

# 1. Learning Objective

A choropleth map is a thematic map composed of **coloured polygons** and is used to represent statistical data using the color mapping symbology technique.

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

-   Plot choropleth maps using tmap package

# 2. Import Packages

```{r}
pacman::p_load(tidyverse, sf, tmap, psych)
```

# 3. Load Data

:::panel-tabset

## Master Plan Subzone

`mpsz` is the Master Plan 2014 Subzone Boundary in ESRI shapefile format. It can be downloaded at data.gov.sg This is a geospatial data. It consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2014. `st_read()` function of sf package to import the MP14_SUBZONE_WEB_PL shapefile into R as a simple feature data frame.

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```
```{r}
mpsz
```

## Population

`pop` is the Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format (i.e. respopagesextod2011to2020.csv). This is an aspatial data fie. It can be downloaded at Department of Statistics, Singapore Although it does not contain any coordinates values, but it’s PA and SZ fields can be used as unique identifiers to geocode to MP14_SUBZONE_WEB_PL shapefile.

```{r}
pop <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
glimpse(pop)
```
:::

# 4. Prepare Data

:::panel-tabset
## Recode

The following attributes will be derived:

YOUNG: Age group `0_to_4` and `20_to_24`
ECONOMY ACTIVE: Age group `25_to_29` and `60_to_64`
AGED: Age group `65_and_above`
TOTAL: All age group
DEPENDENCY: Ratio between YOUNG and AGED : ECONOMIC ACTIVE

```{r}
pop2020 <- pop %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(POP = sum(Pop)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>%
  mutate(ECONOMY_ACTIVE = rowSums(.[7:11]) + rowSums(.[13:15]))%>%
  mutate(AGED = rowSums(.[16:21])) %>%
  mutate(TOTAL = rowSums(.[3:21])) %>%
  mutate(DEPENDENCY = (YOUNG + AGED)/ECONOMY_ACTIVE) %>%
  select(PA, SZ, YOUNG, ECONOMY_ACTIVE, AGED, TOTAL, DEPENDENCY)
glimpse(pop2020)
```
```{r}
describe(pop2020)
```

## Uppercase

`PA` and `SZ` of `pop2020` and `SUBZONE_N` and `PLN_AREA_N` of `mpsz`are standardized to uppercase.

```{r}
pop2020 <- pop2020 %>%
  mutate_at(.vars = vars(PA, SZ), .funs = funs(toupper)) %>%
  filter(ECONOMY_ACTIVE > 0)
describe(pop2020)
```
## Left Join

`pop2020` and `mpsz` are joined using `SZ` and `SUBZONE_N`.

```{r}
mpszpop2020 <- left_join(mpsz, pop2020,
                         by = c("SUBZONE_N" = "SZ"))
glimpse(mpszpop2020)
```
:::

::: {.callout-note collapse="true"}
# Save Point

```{r}
write_rds(mpszpop2020, "data/mpszpop2020.rds")
```
:::


# Choropleth Map

::: {.callout-note collapse="true"}
## Load Point

```{r}
mpszpop2020 <- readRDS("data/mpszpop2020.rds")
```
:::


## qtm()

`qtm()` plots a quick thematic map and is a convenient wrapper of the main plotting method. The `fill` argument is either a colour or a data variable to draw a choropleth. The disadvantge of `qtm()` makes aesthetics of individual layers harder to control.

```{r}
tmap_mode("plot")
qtm(mpszpop2020, fill = "DEPENDENCY")
```
## tmap()

The basic building block of tmap is `tm_shape()` followed by the layer elements `tm_shape()` to define the input data and `tm_polygons()` to draw the planning subzone polygons.

Note that `tm_polygons()` is a wrapper of `tm_fill()` and `tm_border()`. `tm_fill()` shades the polygons by using the default colour scheme and `tm_borders()` adds the borders of the shapefile onto the choropleth map where the `alpha` argument is used to define transparency number between 0 (totally transparent) and 1 (default; not transparent).

tmap provides a total ten data classification methods, namely: fixed, sd, equal, pretty , quantile, kmeans, hclust, bclust, fisher, and jenks. The default interval binning used to draw the choropleth map is called “pretty” and the default colour scheme is YlOrRd of ColorBrewer. By default, Missing value will be shaded in grey.


::: panel-tabset

### pretty

```{r}
tm_shape(mpszpop2020)+
  tm_polygons("DEPENDENCY")
```

a quantile data classification that used 5 classes.

```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

:::



```{r}
tm_shape(mpszpop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

