---
title: "Multivariate Data Analysis"
author: "Wan Kee"
date: "2 February 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  error: true
editor: source
---

# 1. Learning Objectives

To understand complex and multivariate data, each visualization offers a unique way to represent and understand patterns, relationships, and distributions within data that contain more than two variables. They translate numerical data into visual formats, making it easier to detect trends, clusters, outliers, and correlations that might be difficult to discern from raw data alone.

1.  Correlation Matrix

    -   Correlogram: Strength and direction of pairwise relationships
        -   Create correlation matrix using `pairs()` of R Graphics
        -   Plot corrgram using `corrplot` package of R
        -   Create an interactive correlation matrix using `Plotly` R

2.  Multivariate Data Visualization

    -   Ternary plot: Three-dimensional analysis
        -   Build static ternary plots using `ggtern()` function of `ggtern` package and interactive ternary plots using `plot-ly()` function of `Plotly` package.

    -   Glyphs or Star Plot

    -   Heatmap: Structure within complex datasets
        -   Plot static and interactive heatmaps using `heatmaply` package

    -   Parallel coordinate plot
        -   Plot statistic parallel coordinates plots by using `ggparcoord()` of `GGally` package and interactive parallel coordinates plots by using `parcoords` package and `parallelPlot` package.

    -   Parallel plot

3.  Space-constrained visualization of hierarchies

    -   Treemaps: Hierarchical structures
        -   Manipulate data into a treemap structure using `dplyr` package
        -   Plot static treemap using `treemap` package and design interactive treemap using `d3treeR` package

    -   Sunburst diagram

ggtern is a package that extends the functionality of ggplot2, giving the capability to plot ternary diagrams for (subset of) the ggplot2 proto geometries. Ternary diagrams are used frequently in a number of disciplines to graph compositional features for mixtures of three different elements or compounds. It is possible to represent a coordinate system having three (3) degrees of freedom, in 2D space, since the third dimention is linear and depends only on the other two.

# 2. Correlation Matrix

A **correlogram** is a visual representation of the **correlation matrix**, which displays the **correlation coefficients** between a set of variables to assess the **strength** and **direction** of the **pairwise linear relationships** between two variables.

Correlation coefficient measures the type and strength of the relationship between two variables and ranges between -1.0 and 1.0.

-   Perfect linear relationship between the two variables = 1
-   Perfect inverse relationship between the two variables = -1
-   No linear relationship between the two variables = 0

There are three broad reasons for computing a correlation matrix.

-   Reveal the pairwise relationship between high-dimensional variables
-   Input into other analyses to exclude missing values by pairwise relationship
-   Function as model diagnostic to assess multicolinearity

## 2.1 Load Packages

The following R packages will be used:

-   `tidyverse` supports data science, analysis and communication task
-   `corrplot` enables multivariate data visualisation and analysis
-   `ggstatsplot` is an ggplot2 extension for creating graphics with details from statistical tests

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse, ggplot2)
```

## 2.2 Import Data

`wine` is a wine quality dataset with attributes tested in wine chemistry and are used to determine the quality of wine through sensory and chemical analysis.

```{r}
wine <- read_csv("data/wine_quality.csv")
glimpse(wine)
```

::: {.callout-tip collapse="true"}
### Contextual Knowledge on Wine

Attributes of Wine Quality:

Fixed Acidity: This refers to the concentration of tartaric, malic, and other acids that do not evaporate readily. Higher levels can make a wine taste tart or sour.

Volatile Acidity: This measures the acidic compounds in wine that can vaporize easily. At low levels, they can add to the complexity of the wine, but high levels can lead to an unpleasant, vinegar-like aroma.

Citric Acid: As one of the acids found naturally in grapes, it's less common in wines but can be added for flavor and freshness.

Residual Sugar: The amount of sugar remaining after fermentation stops. It affects the sweetness of the wine; dry wines have low residual sugar, while sweet wines have high levels.

Chlorides: The amount of salt in the wine, which can influence flavor.

Free Sulfur Dioxide (SO2): This acts as a preservative by preventing microbial growth and oxidation. It can also affect the smell and taste of the wine.

Total Sulfur Dioxide: The sum of bound and free sulfur dioxide. High levels can lead to off-putting aromas and can affect those sensitive to sulfites.

Density: This can give an indication of the alcohol content and sweetness of the wine. Denser wines might have higher sugar content and potentially higher alcohol levels.

pH: The level of acidity in the wine. Wines with lower pH (more acidic) tend to have a longer shelf life and more tartness.

Sulphates: These are added to wine to preserve it and can influence its fermentation and flavor.

Alcohol: The alcohol level can affect the body, flavor, and texture of the wine.

Quality: This is often a score given by experts based on the balance of the wine's characteristics.

Type: Refers to the classification of the wine (e.g., red, white, rosé) and can be used to correlate specific attributes with the preferred profiles of each type.
:::

## 2.3 Plot Correlation matrix

Correlation matrix can be created using the following packages:

::: panel-tabset
### R Graphics

#### R Graphics

`pair()` from R Graphics produce a matrix of scatterplots given `x` as the numeric columns of a matrix.

Among the 13 attributes in `wine`, the `type` attribute is a categorical variable to be excluded from the matrix.

#### Correlation Coefficient

`panel.cor` function can show the correlation coefficient of each pair of variables.

```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y, use = "complete.obs"))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste(prefix, txt, sep="")
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
  }

pairs(wine[,1:12], 
      upper.panel = panel.cor)
```

::: {.callout-note icon="false"}
##### Observations

Below and above the diagonal are scatter plots for each pair of variables, which show how the variables relate to one another. Tight clustering of points along a line indicates a strong correlation, either positive (sloping upwards) or negative (sloping downwards).

However, given the static nature of the matrix, precise quantitative analysis and correlation coefficients are not discernible.
:::

#### Double Panel

```{r}
pairs(wine[,1:12])
```

::: {.callout-note icon="false"}
##### Observations

Below and above the diagonal are scatter plots for each pair of variables, which show how the variables relate to one another. Tight clustering of points along a line indicates a strong correlation, either positive (sloping upwards) or negative (sloping downwards).

However, given the static nature of the matrix, precise quantitative analysis and correlation coefficients are not discernible.
:::

#### Panel Selection

A correlation matrix is symmetric. `lower.panel` and `upper.panel` are panel functions used below and above the diagonal respectively.

```{r}
pairs(wine[,1:12], upper.panel = NULL)
```

### ggstatplot

#### Ggstatplot Package

`ggcormat()` of ggstatplot package creates a corrgram where `cor.var` produce a correlation coefficient matrix and the associated p-value matrix

```{r}
ggcorrmat(
  data = wine, 
  cor.vars = 1:12,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)
```

#### Customization

```{r}
ggcorrmat(
  data = wine, ## data from which variable is to be taken
  cor.vars = 1:12,
  type = "np", ## which correlation coefficient is to be computed
  lab.col = "red", ## label color
  ggtheme = theme_light(), ## selected ggplot2 theme
  ## turn off default ggestatsplot theme overlay
  matrix.type = "lower", ## correlation matrix structure
  colors = NULL, ## turning off manual specification of colors
  palette = "category10_d3", ## choosing a color palette
  package = "ggsci", ## package to which color palette belongs
  title = "Gapminder correlation matrix", ## custom title
  subtitle = "Source: Gapminder Foundation" ## custom subtitle
)
```

#### Multiple Plots

ggstatsplot provides a special helper function to do separate analysis using `grouped_ggcorrmat()`. It applies `ggcorrmat()` across all levels of a specified grouping variable and then combines list of individual plots into a single plot.

```{r}
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```

### corrplot

#### Corrplot Package

`cor()` of R Stats is used to compute the correlation matrix of wine data frame.

```{r}
wine.cor <- cor(wine[, 1:12])
```

```{r}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

### ellipse

#### Ellipse Package

The ellipse package allows us to build a correlogram using `plotcorr()` function.

First of all, you have to compute the correlation matrix of your dataset using the `cor()` function of R. Each correliation will be represented as an ellipse by the `plotcorr()` function. Color, shape and orientation depend on the correlation value.

```{r}
# Libraries
library(ellipse)
library(RColorBrewer)
 
# Use of the mtcars data proposed by R
wine.cor <- cor(wine[, 1:12]) 

# Build a Pannel of 100 colors with Rcolor Brewer
my_colors <- brewer.pal(5, "Spectral")
my_colors <- colorRampPalette(my_colors)(100)
 
# Order the correlation matrix
ord <- order(wine.cor[1, ])
wine_ord <- wine.cor[ord, ord]
plotcorr(wine_ord , col=my_colors[wine_ord*50+50] , mar=c(1,1,1,1)  )
```

### corrgram

#### Corrgram Package

Using `corrgram()`, there are two ways to graphically display the correlation matrix using the `panel.shade()` and `panel.pie()` functions.

```{r}
library(corrgram)

corrgram(wine[, 1:12], order=TRUE,
         main="Wine",
         lower.panel = panel.shade, upper.panel = panel.pie,
         diag.panel = panel.minmax, text.panel = panel.txt)
```
:::

# 3. Ternary Plot

A **ternary plot**, also known as a ternary graph, triangle plot, simplex plot, or de Finetti diagram, is a barycentric plot on **three variables** which **sum to a constant**. It graphically depicts the ratios of the three variables as points in an equilateral triangle. It is used when the sum of the three variables is fixed, typically at 100 percent, making it useful for representing compositional data where the components of a whole must add up to a constant value.

## 3.1 Load Packages

The following R packages will be used:

-   `tidyverse` supports data science, analysis and communication task

-   `plotly`

-   `ggtern` plots static ternary plots.

-   `Ternary` creates of ternary plots in the R graphics

```{r}
pacman::p_load(tidyverse, plotly, ggtern, Ternary)
```

## 3.2 Import and Prepare Data

`pop` is a population dataset of Singapore residents with attributes on planning area subzone, age group, sex and type of dwelling collected between June 2000-2018.

```{r}
pop <- read_csv("data/respopagsex2000to2018_tidy.csv")
glimpse(pop)
```

We derive **three new measures**, namely **young**, **active**, and **old**, using `mutate()` of dplyr package.

```{r}
pop3 <- pop %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
glimpse(pop)
```

## 3.3 Plot ternary plot

::: panel-tabset
### ggtern

`ggtern()` function of ggtern package to create a simple ternary plot.

```{r}
# Building the static ternary plot
ggtern(data = pop3, aes(x = YOUNG,y = ACTIVE, z = OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

`plot_ly()` function of Plotly R creates an interactive ternary plot.

```{r}
# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  pop3, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

### Ternary

```{r}
# Configure plotting area
par(mar = rep(0.3, 4))

# Example data object
dat <- data.frame(sio2 = c(2, 4, 10, 20),
                  fe2o3 = c(5, 6, 7, 8),
                  al2o3 = c(12, 11, 10, 9),
                  grain_size = c(20, 16, 12, 8),
                  reflectance = c(80, 63, 51, 20))

# Define a colour spectrum
spectrumBins <- 255 # Number of bins to use
mySpectrum <- viridisLite::viridis(spectrumBins)

# Cut our reflectance data into categories
binnedReflectance <- cut(dat$reflectance, spectrumBins)

# Assign each data point a colour from the spectrum
pointCol <- mySpectrum[binnedReflectance]


# Define a size range
maxSize <- 2.4 # Size of largest point, in plotting units
sizes <- dat$grain_size
pointSize <- sizes * maxSize / max(sizes)

# Initialize the plot
TernaryPlot(atip = expression(SiO[2]),
            btip = expression(paste(Fe[2], O[3], " (wt%)")),
            ctip = expression(paste(Al[2], O[3]))
)

# Plot the points
TernaryPoints(dat[, c("sio2", "fe2o3", "al2o3")],
              cex = pointSize, # Point size
              col = pointCol,  # Point colour
              pch = 16         # Plotting symbol (16 = filled circle)
              )

# Legend for colour scale
PlotTools::SpectrumLegend(
  "topleft",
  cex = 0.8, # Font size
  palette = mySpectrum,
  legend = paste(
    seq(from = max(dat$reflectance), to = min(dat$reflectance),
        length.out = 5),
    "%"
  ),
  bty = "n", # No framing box
  xpd = NA, # Don't clip at margins
  # title.font = 2, # Bold. Supported from R 3.6 onwards
  title = "Reflectance"
)

# Legend for point size
PlotTools::SizeLegend(
  "topright",
  width = c(0, maxSize),
  lend = "round", # Round end of scale bar
  legend = paste(
    signif(seq(max(sizes), 0, length.out = 5), digits = 3),
    "\u03bcm" # µm
  ),
  title = "Grain size",
  # title.font = 2, # Bold. Supported from R 3.6 onwards
  bty = "n", # Do not frame with box
  cex = 0.8
)
```
:::

# 4. Heatmap

A heatmap is a data visualization technique that shows magnitude of a phenomenon as color in two dimensions. The variation in color may be by hue or intensity, giving visual cues to the reader about how the phenomenon is clustered or varies over space.

## 4.1 Load Packages

The following R packages will be used:

-   `tidyverse` supports data science, analysis and communication task

-   `seriation`

-   `dendextend`

-   `heatmaply`

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## 4.2 Import and Prepare Data

`happiness` is a world happines report from 2018.

```{r}
happiness <- read_csv("data/WHData-2018.csv")
glimpse(happiness)
```

change the rows by country name instead of row number

```{r}
row.names(happiness) <- happiness$Country
```

```{r}
happiness1 <- dplyr::select(happiness, c(3, 7:12))
happiness_matrix <- data.matrix(happiness)
```

## 4.3 Plot Heatmap

::: panel-tabset
### Base Stats

`heatmap()` plots a cluster heatmap by reordering using clusterisation; it calculates the distance between each pair of rows and columns and try to order them by similarity. The corresponding **dendrogram** are provided beside the heatmap.

```{r}
heatmap(happiness_matrix)
```

The `scale` argument normalize the matrix. The `margins` argument is used to ensure that the entire x-axis labels are displayed completely and, `cexRow` and `cexCol` arguments are used to define the font size used for y-axis and x-axis labels respectively.

The arguments `Rowv = NA` and `Colv = NA` are used to switch off the option of plotting the row and column dendrograms.

```{r}
heatmap(happiness_matrix,
        scale="column",
        cexRow = 0.6, 
        cexCol = 0.8,
        margins = c(10, 4))

# Switch off dendrogram
# heatmap(happiness_matrix, Rowv = NA, Colv = NA)
```

### heatmaply

heatmaply is an R package for building interactive cluster heatmap using `heatmaply()` and it can be shared online as a stand-alone HTML file.

```{r}
heatmaply(happiness_matrix[, -c(1, 2, 4, 5)])
```
:::

## 4.4 Data Transformation within heatmaply()

Three main data transformation methods are supported by heatmaply(), namely: scale, normalise and percentilse.

::: panel-tabset
### Scale

Assume the variables have a normal distribution, scaling will bring the values close to the standard normal distribution where each value reflects the distance from the mean in units of standard deviation.

Scaling subtract the mean from each values and divide by the standard deviation.

The `scale` argument in `heatmaply()` supports column and row scaling.

```{r}
heatmaply(happiness_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

### Normalise

When variables in the data comes from possibly different (and non-normal) distributions, the normalize function can be used to bring data to the 0 to 1 scale by subtracting the minimum and dividing by the maximum of all observations.

This preserves the shape of each variable’s distribution while making them easily comparable on the same “scale”.

```{r}
heatmaply(normalize(happiness_matrix[, -c(1, 2, 4, 5)]))
```

### Percentalise

This is similar to ranking the variables, but instead of keeping the rank values, divide them by the maximal rank. This is done by using the ecdf of the variables on their own values, bringing each value to its empirical percentile. The benefit of the percentize function is that each value has a relatively clear interpretation, it is the percent of observations that got that value or below it.

```{r}
heatmaply(percentize(happiness_matrix[, -c(1, 2, 4, 5)]))
```
:::

## 4.5 Clustering Algorithm

Heatmaply supports a variety of hierarchical clustering algorithm. The main arguments provided are:

-   distfun: function used to compute the distance (dissimilarity) between both rows and columns. Defaults to dist. The options “pearson”, “spearman” and “kendall” can be used to use correlation-based clustering, which uses as.dist(1 - cor(t(x))) as the distance metric (using the specified correlation method).

-   hclustfun: function used to compute the hierarchical clustering when Rowv or Colv are not dendrograms. Defaults to hclust.

-   `dist_method` default is NULL, which results in “euclidean” to be used. It can accept alternative character strings indicating the method to be passed to distfun. By default distfun is “dist”” hence this can be one of “euclidean”, “maximum”, “manhattan”, “canberra”, “binary” or “minkowski”.

-   `hclust_method` default is NULL, which results in “complete” method to be used. It can accept alternative character strings indicating the method to be passed to hclustfun. By default hclustfun is hclust hence this can be one of “ward.D”, “ward.D2”, “single”, “complete”, “average” (= UPGMA), “mcquitty” (= WPGMA), “median” (= WPGMC) or “centroid” (= UPGMC).

To determine the best clustering method and the number of cluster, the `dendextend` package will be used.

`dend_expend()` will be used to determine the recommended clustering method to be used.

```{r}
wh_d <- dist(normalize(happiness_matrix[, -c(1, 2, 4, 5)]), 
             method = "euclidean")
dend_expend(wh_d)[[3]]
```

::: {.callout-note appearance="simple"}
### Observation

The “average” method should be used because it gave the high optimum value, while "single" has the lowest value.
:::

`find_k()` is used to determine the optimal number of cluster.

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

::: {.callout-note appearance="simple"}
### Observation

The optimal number of cluster is 3.
:::

```{r}
heatmaply(normalize(happiness_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

## 4.6 Seriation

Seriation in the context of a heatmap refers to the process of rearranging the rows and columns of the data matrix such that similar values are placed near each other. This is typically done to reveal patterns, such as clusters or gradients, more clearly.

::: panel-tabset
### OLO

```{r}
heatmaply(normalize(happiness_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

### GW

```{r}
heatmaply(normalize(happiness_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

### mean

```{r}
heatmaply(normalize(happiness_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

### none

```{r}
heatmaply(normalize(happiness_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```
:::

## 4.7 Customization

The following arguments are used:

-   `k_row` is used to produce 5 groups

-   `margins` is used to change the top margin to 60 and row margin to 200

-   `fontsizw_row` and `fontsize_col` are used to change the font size for row and column labels to 4

-   `main` is used to write the main title of the plot

-   `xlab` and `ylab` are used to write the x-axis and y-axis labels

```{r}
heatmaply(normalize(happiness_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```

# 5. Parallel Coordinates Plot

A **parallel coordinates plot** is a type of visualization that allows one to see the patterns in multidimensional data. It is particularly useful for visualizing high-dimensional geometry and analyzing multivariate data.

## 5.1 Load Packages

The following R packages will be used:

-   `tidyverse` supports data science, analysis and communication task

-   `parallelPlot`

-   `GGally`

-   `parcoords`

```{r}
pacman::p_load(tidyverse, parallelPlot, GGally, parcoods)
```

## 5.2 Import and Prepare Data

`happiness` is a world happines report from 2018.

```{r}
happiness <- read_csv("data/WHData-2018.csv")
glimpse(happiness)
```

## 5.3 Plot Parallel Coordinates Plot

`ggparcoord()` plot a basic static parallel coordinates plot where `data` argument is used to map the data object, `columns` is used to select the columns for preparing the parallel coordinates plot.

::: panel-tabset
### Highlight

`splineFactor` is a logical or numeric operator indicating whether spline interpolation should be used. `missing` is a method used to handle missing values.

To rotate x-axis text labels, we use `axis.text.x` as argument to `theme()` function. And we specify `element_text(angle = 30)` to rotate the x-axis text by an angle 30 degree.

Data visualization aims to highlight a story in the data. If you are interested in a specific group, you can highlight it using `scale_color_manual()`.

```{r}
happiness$Highlight <- ifelse(happiness$Country == "Singapore", "Singapore", "Other")

ggparcoord(data = happiness,
           columns = c(7:12), 
           groupColumn = "Highlight",
           missing = "exclude",
           splineFactor = TRUE) +
  scale_color_manual(values = c("Singapore" = "hotpink", "Other" = "grey")) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30))
```

### Boxplot

Boxplots can be added to parallel coordinates plots to provide a summary of the distribution of data for each variable. The center Line of the boxplot represents the median of the data for that particular variable.

`groupColumn` argument is used to group the observations by using a single variable and colour the parallel coordinates lines by region name.

`scale` argument is used to scale the variables in the parallel coordinate plot by using uniminmax method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one.

`alphaLines` argument is used to reduce the intensity of the line colour to 0.2. The permissible value range is between 0 to 1.

`boxplot` argument is used to turn on the boxplot by using logical TRUE. The default is FALSE. title argument is used to provide the parallel coordinates plot a title.

```{r}
ggparcoord(data = happiness, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 30))
```

### Facet

`facet_wrap()` of ggplot2 is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}
ggparcoord(data = happiness, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 30))
```

### Interactive

`parallelPlot()` plot an interactive parallel coordinates plot. `rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
wh <- happiness %>%
  select("Happiness score", c(7:12))

parallelPlot(wh,
             width = 300,
             height = 500,
             rotateTitle = TRUE)
```

`histoVisibility` argument is used to plot histogram along the axis of each variables.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```
:::

# 6. Treemap

A **treemap** is a data visualization technique used to display **hierarchical** (tree-structured) data using nested rectangles. Each branch of the tree is represented by a rectangle, which is then tiled with smaller rectangles representing sub-branches. A leaf node's rectangle has an area proportional to a specified dimension of the data. Often, the leaf nodes are colored to show a separate dimension of the data.

## 6.1 Load Packages

The following R packages will be used:

-   `tidyverse` supports data science, analysis and communication task

-   `treemapify`

-   `treemap`

```{r}
pacman::p_load(tidyverse, treemapify, treemap, d3Tree)
```

## 6.2 Import and Prepare Data

`realis` provides information of private property transaction records in 2018.

```{r}
realis <- read_csv("data/realis2018.csv")
glimpse(realis)
```

## Transform Data

::: panel-tabset
### Summary

```{r}
realis_summarised <- realis %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
glimpse(realis_summarised)
```

### Selection

```{r}
realis_selected <- realis_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
glimpse(realis_selected)
```
:::

## 6.3 Plot Treemap

::: panel-tabset
### Treemap

Treemap by using three core arguments of treemap(), namely: `index`, `vSize` and `vColor`.

```{r}
treemap(realis_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### geom_treemap

`geom_treemap()` from ggplot2 plackage to draw a tree map. `geom_treemap()` requires an `area` aesthetic and does not interpret x and y; axes are not meaningful in a treemap.

```{r}
ggplot(data=realis_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") + 
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

### geom_treemap with hierarchy

An optional `subgroup` aesthetic will cause the tiles to be clustered in subgroups within the treemap. Using the `subgroup` and `subgroup2` arguments, we will visualize the treemap with planning area and planning region. `subgroup3` is available where visualization requires.

To add text labels to tiles, use `geom_treemap_text()`.

```{r}
ggplot(data = realis_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "white")
```

### d3tree

**d3tree** package creates flexible interactive visualization for large trees using D3.js. The structure of a D3Tree is specified with the `index` and each node stored in a vector.

```{r}
tm <- treemap(realis_summarised,
              index=c("Planning Region", "Planning Area"),
              vSize="Total Unit Sold",
              vColor="Median Unit Price ($ psm)",
              type="value",
              title="Private Residential Property Sold, 2017",
              title.legend = "Median Unit Price (S$ per sq. m)")
```

`d3tree()` is the primary function for creating interactive d3.js treemaps from various data types. The `data` argument is the data to be plotted, `id` is the default to allow nodes with non-unique names.

```{r}
d3tree(tm, rootname = "Singapore")
```
:::

# 7. Reference

ggtern http://www.ggtern.com/d/2.2.0/

ggtern github https://github.com/cran/ggtern

ellipse: https://r-graph-gallery.com/97-correlation-ellipses.html#:\~:text=The%20ellipse%20package%20allows%20to,by%20the%20plotcorr()%20function.

corrgram: https://cran.r-project.org/web/packages/corrgram/vignettes/corrgram_examples.html
